#Lab_03 {lab_03}

rob wells, 8/14/2021

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries and establish settings
```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse here


# Load janitor here


```

### Load data

In this data set, each row represents a city or town in Maryland with the 2010 and 2020 figures for: 

-  Total population 
-  White only population
-  Black or African American only population
-  Hispanic only population 
-  Native American only population
-  Asian only population
-  Pacific Islander (including native Hawiian) only population

Background on this data from [Social Explorer](https://www.socialexplorer.com/data/ACS2016_5yr/metadata/?ds=ACS16_5yr&table=B02001) 
```{r}
# Write code to load the Maryland race by city, place for 2010 - 2020 and call it md_race.
# The data is called lab_03.csv, and is in the same folder as lab_03.qmd (this data notebook)

# your answer here

MarylandData <- rio::import("lab_03.csv" , which = "RaceData")

View(MarylandData)


```


## Answer questions

Q1: Which five cities have the largest overall Hispanic population in 2020? Name the five cities that had the largest Hispanic population in 2020. 
The five cities that had the largest Hispanic population in 2020 are Baltimore City, Wheaton CDP, Germantown CDP, Silver Spring CDP, and Aspen Hill CDP. 

A1:
```{r}

# Write your code in this codeblock.
hispanic <- MarylandData %>% 
  select( place, x2020_hispanic) %>%
  slice_max(x2020_hispanic, n=5)
```

Q2: Find the cities with the largest Black population and arrange the results in descending order. 
A2: Baltimore City had the largest Black population, see page titled "Black" for more. 

```{r}
#write your code here
black <- MarylandData %>% 
  select( place, x2020_black) %>%
  arrange(desc(x2020_black))

# your answer here

```

Q3: Create a table called top_hispanic that shows the top 15 communities with the largest Hispanic population in 2020, and also display the 2010 Hispanic population and the total community population for 2020. Look at your results - what stands out to you? Craft a one-paragraph story pitch for your editor. 

A3: Overall, the Hispanic population has increased in the state of Maryland. One thing I would look into is the increase in Frederick City. Frederick City increased the Hispanic population from 7,130 to 11,937, which is over a 50% increase. I would look into if there was a reason that drew Hispanics to this area, or if there is a reason why there was an increase. 

```{r}

# your answer here
hispanic <- MarylandData %>% 
  select( place, x2020_hispanic, x2010_hispanic, x2020_total) %>%
  slice_max(x2020_hispanic, n=15)
Me
```
