---
title: "Untitled"
format: html
editor: visual
  markdown: 
    wrap: 72
---

#RSW comment - Becca - You want to have a yaml header (YAML = yet another markup language) on your quarto documents. I pasted one here. I don't have them on my documents because I made a book from the pages, which is another story altogether.

#Milestone_04_Final_Project

##Background

The purpose of this project is to analyze the temperature of Lake Michigan~~. The data analysis will focus on the comparison to the last full year (2023), but also analyze~~ from 2023 ~~in relation to past years~~. XXX whenXXX Lastly, it will analyze Lake Michigan's temperature to the temperature of the other four Great Lakes.

#Load Libraries

```{r}
library(tidyverse)
library(tidycensus)
library(rvest)
library(janitor)
library(dplyr)
library(readxl)
library(stats)
library(ggplot2)
library(lubridate)
library(tidyr)
```

#RSW comment : format this so the questions are on separate lines rather than all bunched together

#Questions 1. In 2023, were there noticeable differences in the temperature of Lake Michigan compared to temperature differences in the other four great lakes? 2. What 5 days did Lake Michigan have the biggest drop in temperature in 2023? What 5 days did Lake Michigan have the biggest increase in temperature in 2023? 3.How have the average temperatures of Lake Michigan changed over the past five years, and are there regional weather patterns or recreational activities that could have impacted these changes? 4. How does 2023's temperature patterns and changes compare to the patterns of those one and two decades ago (2013 and 2003)? 5. How does Lake Michigan's temperature compare to the average trends in land temperature in Michigan throughout the 2023 year?

#Load Data from last 5 years

```{r}


data23 <- rio::import("temp23.txt") %>% 
  clean_names() %>% 
  rename(year= "v1", day="v2", Sup.="v3", Mich.="v4", Huron="v5",Erie="v6", Ont.="v7",St.Clr="v8") %>% 
clean_names()

data22 <- rio::import("temp2022.txt") %>% 
  clean_names() %>% 
  rename(year= "v1", day="v2", Sup.="v3", Mich.="v4", Huron="v5",Erie="v6", Ont.="v7",St.Clr="v8") %>% 
clean_names()

data21 <- rio::import("temp2021.txt") %>% 
  clean_names() %>% 
  rename(year= "v1", day="v2", Sup.="v3", Mich.="v4", Huron="v5",Erie="v6", Ont.="v7",St.Clr="v8") %>% 
clean_names()

data20 <- rio::import("temp2020.txt") %>% 
  clean_names() %>% 
  rename(year= "v1", day="v2", Sup.="v3", Mich.="v4", Huron="v5",Erie="v6", Ont.="v7",St.Clr="v8") %>% 
clean_names()

data19 <- rio::import("temp2019.txt") %>% 
  clean_names() %>% 
  rename(year= "v1", day="v2", Sup.="v3", Mich.="v4", Huron="v5",Erie="v6", Ont.="v7",St.Clr="v8") %>% 
  clean_names
```

#Load Data from 2013 and 2003

```{r}
data13 <- rio::import("temp2013.txt") %>% 
  clean_names() %>% 
  rename(year= "v1", day="v2", Sup.="v3", Mich.="v4", Huron="v5",Erie="v6", Ont.="v7",St.Clr="v8") %>% 
  clean_names()

data03 <- rio::import("temp2003.txt") %>% 
  clean_names() %>% 
  rename(year= "v1", day="v2", Sup.="v3", Mich.="v4", Huron="v5",Erie="v6", Ont.="v7",St.Clr="v8") %>% 
clean_names()

```

#Load Average Temperatures of 2023
#RSW comment - these are the average air temperatures in fahrenheit?

```{r}
manis_temp <- read_excel("manis_temp.xltx")
```

#Question 1 In 2023, were there noticeable differences in the average temperature of Lake Michigan compared to temperature differences in the other four great lakes?

```{r}

#RSW comment - delete this - it's old
# mich_avg_2023 <- data23 
#  mean(data23$sup)
#  mean(data23$mich)
#  mean(data23$huron)
#  mean(data23$erie)
#  mean(data23$ont)
#  mean(data23$st_clr)

#Mutate to own row 
 
mich_avg_2023 <- data23 %>% 
mutate(sup = mean(data23$sup), 
 mich = mean(data23$mich),
 huron = mean(data23$huron),
 erie = mean(data23$erie),
 ont =  mean(data23$ont),
st_clr = mean(data23$st_clr)) %>% 
  head(1) %>% 
  select(sup, mich, huron, erie, ont, st_clr)
```

#Observations Lake Michigan was recorded on average lower temperatures (10.53 degrees Celsius average ) than every lake except Lakes Huron (9.80 Celsius) and Superior(6.89 Celsius). Lake Superior is North of Lake Michigan, so expected to be at a cooler temperature. Lake Huron is North-East of Lake Michigan, and isn't as large, so most likely cools faster.

#Link to Data https://coastwatch.glerl.noaa.gov/statistics/average-surface-water-temperature-glsea/ -\> Home Page https://apps.glerl.noaa.gov/coastwatch/ftp/glsea/avgtemps/2023/glsea-temps2023_1024_3.dat -\>Set Used

#Question 2 What 5 days did Lake Michigan have the biggest changes in temperature of 2023?

#Rsw comment: this table needs the months added to it

```{r}
mich_droplag <- data23 %>% 
  select(year, day, mich) %>%
  arrange(day) %>%
  mutate(temp_diff = mich - lag(mich, default = first(mich))) %>% 
    arrange(desc(temp_diff))
```

#Observations The 5 biggest changes in temperature for 2023 were all increases in temperature. The largest change was day 149, increasing by 1.09 degrees. This was followed by day 150 at 1.06 degrees, day 151 at 0.89 degrees, day 153 at 0.80 degrees. The fifth biggest increase was on day 152, 154 and 182, which they all increased by 0.69 degrees. The biggest decrease was on day 158, by -0.58. Days 150-160 of 2023 ranged from May 30th to June 8th. This is when the temperatures start to increase in the air by a significant amount as Summer approached, so it is reasonable to expect a correlation to how the temperatures of the water are changing. This code selects the year 2023, all 365 days and solely selects Lake Michigan, which the question is focusing on. Then, it figures out the difference between one row to the next. I arranged it by day and the difference in temperature, so it would know to arrange the difference in rows based on the difference in temperature (temp_diff) by day. Then, I arranged it descending, so that the biggest increases would appear at the top, and biggest decreases appeared at the bottom. However, the top 5 biggest changes were all increases.

#Link to Data https://coastwatch.glerl.noaa.gov/statistics/average-surface-water-temperature-glsea/ -\> Home Page https://apps.glerl.noaa.gov/coastwatch/ftp/glsea/avgtemps/2023/glsea-temps2023_1024_3.dat -\>Set Used

#Question 3 How have the average temperatures of Lake Michigan changed over the past five years, and are there regional weather patterns or recreational activities that could have impacted these changes?

```{r}
#Calculate the average Temperature for each year, and get rid of 2024
year_avg <- bind_rows(data19, data20, data21, data22, data23) %>%
  group_by(year) %>%
  summarise(year_avg = mean(mich))
#Make a plot
ggplot(year_avg, aes(x = year, y = year_avg)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Average Temperature of Lake Michigan", title = "Lake Michigan's Average Temperature Trends")

```

#Observations There was an upwards trend in Lake Michigan's temperature from 2019 to 2021, with a decrease in 2022. However, the rise in temperatures in 2023 was significant. Just the difference between 2022 and 2023 rose from about 9.7 degrees Celsius to approximately 10.6 degrees Celsius.

#Data Used #Link to Data https://coastwatch.glerl.noaa.gov/statistics/average-surface-water-temperature-glsea/ -\> Home Page https://apps.glerl.noaa.gov/coastwatch/ftp/glsea/avgtemps/2023/glsea-temps2023_1024_3.dat https://apps.glerl.noaa.gov/coastwatch/ftp/glsea/avgtemps/2022/glsea-temps2022_1024.dat https://apps.glerl.noaa.gov/coastwatch/ftp/glsea/avgtemps/2021/glsea-temps2021_1024.dat https://apps.glerl.noaa.gov/coastwatch/ftp/glsea/avgtemps/2020/glsea-temps2020_1024.dat https://apps.glerl.noaa.gov/coastwatch/ftp/glsea/avgtemps/2019/glsea-temps2019_1024.dat -\>Sets Used

#Question 4 How does 2023's temperature patterns and changes compare to the patterns of those one and two decades ago (2013 and 2003)?

#RSW comment: the x axis needs to be in months, not days of the year

```{r}
# Select data
data23_select <- select(data23, year, day, mich)
data13_select <- select(data13, year, day, mich)
data03_select <- select(data03, year, day, mich)

# Filter data for 365 days of each year (since none were leap)
data23_all_days <- filter(data23_select, day <= 365)
data13_all_days <- filter(data13_select, day <= 365)
data03_all_days <- filter(data03_select, day <= 365)

# Plot
ggplot() +
  geom_line(data = data23_all_days, aes(x = day, y = mich, color = "2023")) +
  geom_line(data = data13_all_days, aes(x = day, y = mich, color = "2013")) +
  geom_line(data = data03_all_days, aes(x = day, y = mich, color = "2003")) +
  labs(x = "Day of the year", y = "Temperature (°C)", color = "Year",
       title = "Temperature of Lake Michigan Daily by Decade")
```

#Answer: The temperature of Lake Michigan stayed consistent with trends over the last two decades. However, 2013 had a jump in temperature between approx. day 185 and 200. 2023 has this jump much earlier, from day 145 to day 165. While 2003 did not have a drastic jump in temperature. As displayed, the 2023 temperature line is higher than the other lines towards the end of the year. The 2003 line is also visibly the lowest line for most of the year. This means that over the last two decades, the temperature of Lake Michigan has slowly increased.

#Question 5 How does Lake Michigan's temperature compare to the average trends in land temperature of Manistee, MI. Manistee is on the coast of Lake Michigan. (manis temp) in Michigan throughout the 2023 year?

#RSW comment - move this higher so all of the following data can be analyzed by the specific month

```{r}
#Convert Data23 to months, to match Manistee.
data23 <- data23 %>%
  mutate(month = case_when(
    day %in% 1:31 ~ "Jan",
    day %in% 32:60 ~ "Feb",
    day %in% 61:91 ~ "Mar",
    day %in% 92:121 ~ "Apr",
    day %in% 122:152~ "May",
    day %in% 153:182 ~ "Jun",
    day %in% 183:212 ~ "Jul",
    day %in% 213:243 ~ "Aug",
    day %in% 244:273 ~ "Sep",
    day %in% 274:304 ~ "Oct",
    day %in% 305:333 ~ "Nov",
    day %in% 334:365 ~ "Dec",
  ))

#Find the Monthly average for data23
monthly_avg_mich <- data23 %>%
  group_by(month) %>%
  summarize(avg_temp = mean(mich, na.rm = TRUE))

#Flip Manistee table to horizontal
manis_temp1 <- manis_temp %>% 
pivot_longer(names_to = "month",
             values_to="temp",
             cols=2:13)
```

#RSW comment: Some big problems here. It seems the air temps are in fahrenheit and the water in celcius? So convert the air to celcius so the scale in compatible.




```{r}
#Merge the Data
merged_data <- merge(monthly_avg_mich, manis_temp1, by = "month", all = TRUE) %>% 
mutate(month2=paste(month,"2023",sep=","))
  

#Make a Bar Chart
merged_data$avg_temp <- as.numeric(merged_data$avg_temp)
merged_data$temp <- as.numeric(merged_data$temp) 

merged_data$avg_temp <- round(merged_data$avg_temp, 2)
merged_data$temp <- round(merged_data$temp, 2)

# Plotting
ggplot(merged_data, aes(x = month2, fill=temp)) +
  geom_line(aes(y = avg_temp, color = "Manistee")) +
  geom_line(aes(y = temp, color = "Data23")) +    
  labs(title = "Monthly Average Temperatures: Manistee vs. data23",
       y = "Temperature in Degrees", color = "Dataset") +  
  scale_y_continuous(labels = scales::number_format(accuracy = 0.01))

# RSW comment - just do this
# Convert 'month2' to a factor with ordered levels
merged_data$month2 <- factor(merged_data$month2, levels = c("Jan,2023", "Feb,2023", "Mar,2023", "Apr,2023", "May,2023", "Jun,2023", "Jul,2023", "Aug,2023", "Sep,2023", "Oct,2023", "Nov,2023", "Dec,2023"))

Then create a simple bar chart.

```

#Note to Professor Wells: I still am not able to get this data to run. I cannot get a line to appear, but discovered a further issue. The Manistee data is in Farenheit. So, I went into Excel and made a dataset which was converted to Celsuis for the air temperatures. However, I couldn't get that data to run correctly, so I left the original for now. I will need to come to office hours before class, as I will not be able to get this data to run correctly.

#Converted Data

```{r}
manis_temptest <- read_excel("airtemp.xlsx")
```

#Link to Data https://coastwatch.glerl.noaa.gov/statistics/average-surface-water-temperature-glsea/ -\> Home Page https://apps.glerl.noaa.gov/coastwatch/ftp/glsea/avgtemps/2023/glsea-temps2023_1024_3.dat -\>Set Used https://www.weather.gov/wrh/Climate?wfo=apx - Manistee Temperatures

#Memo From examining the temperature trends of Lake Michigan, there were several significant patterns. First, Lake Michigan consistently had lower average temperatures compared to most other Great Lakes in 2023, with Lakes Huron and Superior recording a slightly lower average. This can be attributed to the geographical location of Lake Michigan, which is further south than Lake Superior and Huron. Additionally, smaller lakes like Lake Huron potentially will cool or warm faster. Furthermore, an upward trend in Lake Michigan's temperature is evident from 2019 to 2021, followed by a slight decrease in 2022. However, 2023 witnessed a significant rise, with temperatures that range from approximately 9.7 to 10.6 degrees Celsius, indicating a significant difference from previous years. Additionally, the analysis from two decades (2003, 2013 and 2023) showed long term insights. While 2013 had a temperature jump between days 185 and 200, 2023 experienced a similar thing much earlier, between days 145 and 165. But, 2003 did not exhibit a drastic temperature fluctuation. The temperature trend lines of the years indicate a consistent increase over the past two decades, with 2023 showcasing the highest overall temperatures. The main finding of this data set so far is that the temperature of Lake Michigan is overall rising over the years. While it is not a consistent rise, the trend is that as time goes on, the temperature of the water is rising. The next step that will add to a significant part of the story, is the comparison of the changes in Lake Michigan’s temperature, and the changes in the air temperature of Manistee, one of the main coastal areas on Lake Michigan’s border.
