---
title: "Project 1"
output: html_document
date: "2024-02-15"
Rebecca Redeker 02/26/2024

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
---
```{r}
library(tidyverse)
library(rio)
library(janitor)
library(ggplot2)
options(scipen=999)

```

#Load Data
In this data set, each row represents the percentage of students that are eligible for free meals from public schools, based on their median household income, in the Baltimore area. 

In this data set, there are rows, referred to as "farmxx" that include the percentages of students who were eligible. The xx stands for the year that the school year finished, ranging from 2010 to 2013. There are also two rows that refer the length and area of the designated town. CSA 210 also stands for the name of the city. 

#Note:
This does not include private instituions within the area.

Background on this data from [Vital Signs Open Data Reporter](https://vital-signs-bniajfi.hub.arcgis.com/datasets/bniajfi::percentage-of-students-receiving-free-or-reduced-meals/about) 

```{r}
#Set working directory to ~/Documents/GitHub/Rebecca_Redeker_JOUR472/Project 1 

meal_fund <- rio::import("meal_fund.xlsx",sheet="Sheet1",skip=1)

```
#Questions
#1: What is the overall distribution of the percentage of students receiving free or reduced meals?
#2:  How does the percentage of students receiving free or reduced meals vary across different areas in Baltimore?
#3: Is there a trend in the percentage of students receiving free or reduced meals over the years?
#4: What is the correlation between the percentage of students receiving free or reduced meals and the area of the neighborhood?


Question 1: What is the overall distribution of the percentage of students receiving free or reduced meals in the Baltimore area in 2010?



#Professor Wells: 
What is Binwidth? R is recommending that I use it, but it does change the way my graph appears.


```{r}

ggplot(data = meal_fund, aes(x = farms10)) +
  geom_histogram(fill = "blue", color = "black") +
  labs(title = "Distribution of Reduced Meals in 2010", x = "Percentage")

```
#Answer #1: 
It appears that many areas within Baltimore had nearly 80% of students receiving free or reduced meals, through public schools. 



#Question 2: How does the percentage of students receiving free or reduced meals vary across different area/size of community in Baltimore for 2013?
```{r}
ggplot(data = meal_fund, aes(x = FID, y = farms13)) +
  geom_boxplot(fill = "orange", color = "black") +
  labs(title = "Reduced meals in different Baltimore areas", x = "Neighborhood", y = "Percentage")
```

#Answer:
Most neighborhoods in 2013 had between 55% and 85% of students receiving free or reduced meals. On average, it was approximately 81% of students.


#Question 3:Is there a trend in the percentage of students receiving free or reduced meals over the years?
```{r}
ggplot(meal_fund, aes(x=FID)) +
  geom_line(aes(y=farms10), color = "blue") +
  geom_line(aes(y=farms11), color = "yellow") +
  geom_line(aes(y=farms12), color = "orange") +
  geom_line(aes(y=farms13), color = "purple") +
  labs(
    title="Trends in Reduced Meals from 2010-2013",
    x = "Area",
    y = "Percentage",
    caption = "source: Vital-Signs Data Portal") 
```

#Answer: 
The biggest change in percentages appeared between 2012 (orange), and 2013 (purple). It would be prominent to dive into if there was a reason, because especially in communities with a very small or very large area, 2013 had the highest percentage, and 2012 had the lowest percentage. 



#Question 4: Question: Is there any correlation between the percentage of students receiving free or reduced meals and the area of the neighborhood, for 2012?
```{r}
ggplot(data = meal_fund, aes(x = SHAPE_Area, y = farms12)) +
  geom_point(color = "blue") +
  labs(title = "Correlation Between Percentage and size of town", x = "Area", y = "Percentage")
```
#Answer:
While there isn't a clear linear pattern, there is correlation. As the area increased, the percentage of students receiving free or reduced meals decreased. However, there are outliers on the far right. 

